---
title: "LRSomatic — per-sample genomics report"
date: today
format:
  html:
    self-contained: true
    toc: true
    toc-depth: 3
    toc-location: left
    theme: flatly
    code-fold: true
    page-layout: full
params:
  sample_id:     "SAMPLE"
  sample_dir:    ""
  reference:     "t2t"
  sex:           "female"
  gene_panel:    "lymphoid"
  default_panel: "lymphoid"
  all_panels:    NULL   # named list from load_all_gene_panels()
  title:         "LRSomatic Report"
  repo_dir:      "."
  outputs:       NULL   # named list from locate_outputs()
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
suppressPackageStartupMessages({
  library(data.table)
  library(dplyr)
  library(DT)
  library(htmltools)
  library(ggplot2)
})

repo_dir = params$repo_dir
source(file.path(repo_dir, "R/utils.R"))
source(file.path(repo_dir, "R/references.R"))
source(file.path(repo_dir, "R/locate_outputs.R"))
source(file.path(repo_dir, "R/parse_smallvariants.R"))
source(file.path(repo_dir, "R/parse_severus.R"))
source(file.path(repo_dir, "R/parse_ascat.R"))
source(file.path(repo_dir, "R/parse_qc.R"))
source(file.path(repo_dir, "R/circos.R"))

outputs = params$outputs
sample_id = params$sample_id

# Load reference data
cytobands    = load_cytobands(params$reference, file.path(repo_dir, "assets"))
chrom_lens   = load_chrom_lengths(params$reference, file.path(repo_dir, "assets"))
chromosomes  = chromosomes_for_sex(params$sex)
chromosomes  = chromosomes[chromosomes %in% unique(cytobands$chrom)]

# Load gene panel
panel_arg = params$gene_panel
panel_genes = tryCatch(
  resolve_gene_panel(panel_arg, file.path(repo_dir, "assets")),
  error = function(e) { message("Gene panel error: ", e$message); character(0) }
)

# Parse ASCAT
ascat_segments = parse_ascat_segments(outputs$ascat_segments)
ascat_pp       = parse_ascat_purityploidy(outputs$ascat_purityploidy)

# Parse QC
qc_mosdepth = parse_mosdepth_summary(outputs$mosdepth_summary)
qc_mosdist  = parse_mosdepth_dist(outputs$mosdepth_dist)
qc_cramino  = parse_cramino(outputs$cramino)
qc_flagstat = parse_flagstat(outputs$flagstat)

# Parse VEP + raw callers
vep_data = parse_vep_text(outputs$vep_somatic)

caller_list = list()
if (!is.null(outputs$clairsto_somatic))
  caller_list$clairsto = parse_caller_vcf(outputs$clairsto_somatic, "clairsto")
if (!is.null(outputs$clairs_somatic))
  caller_list$clairs   = parse_caller_vcf(outputs$clairs_somatic, "clairs")
if (!is.null(outputs$deepsomatic_vcf))
  caller_list$deepsomatic = parse_caller_vcf(outputs$deepsomatic_vcf, "deepsomatic")

variant_table = build_variant_table(vep_data, caller_list, gene_panel = NULL)

# Parse Severus
sev_vcf  = parse_severus_vcf(outputs$severus_vcf)
sev_tsv  = parse_severus_gene_tsv(outputs$severus_gene_tsv)
sv_table = build_sv_table(sev_tsv, gene_panel = NULL)

# SNV data for circos (from VEP file — contains all somatic variants)
snv_circos = NULL
if (!is.null(vep_data) && nrow(vep_data) > 0) {
  snv_circos = unique(vep_data[, .(chrom, pos, ref, alt)])
}

# Draw circos to temp file, then embed as base64
circos_tmp = tempfile(fileext = ".svg")
tryCatch({
  draw_circos(
    snv_data    = snv_circos,
    sv_nontrans = sev_vcf$nontrans,
    sv_trans    = sev_vcf$translocations,
    cnv_data    = ascat_segments,
    cytobands   = cytobands,
    chrom_lengths = chrom_lens,
    chromosomes = chromosomes,
    output_path = circos_tmp
  )
}, error = function(e) {
  message("Circos plot failed: ", e$message)
  circos_tmp <<- NULL
})

# Summary counts
n_snv        = if (!is.null(vep_data)) nrow(unique(vep_data[, .(chrom, pos, ref, alt)])) else NA_integer_
n_sv         = if (!is.null(sev_vcf)) nrow(sev_vcf$nontrans) + nrow(sev_vcf$translocations) else NA_integer_
n_panel_vars = if (!is.null(variant_table) && length(panel_genes) > 0)
                 sum(variant_table$symbol %in% panel_genes, na.rm = TRUE) else 0L
n_panel_svs  = if (!is.null(sv_table) && length(panel_genes) > 0 && "NHL_GENE_HITS" %in% names(sv_table))
                 sum(vapply(sv_table$NHL_GENE_HITS, function(h)
                   any(trimws(unlist(strsplit(as.character(h), "[;,]+"))) %in% panel_genes),
                   logical(1)), na.rm = TRUE) else 0L
```

```{r panel-js-data, results='asis'}
all_p = if (!is.null(params$all_panels) && length(params$all_panels) > 0)
          params$all_panels else list()
js_panels = paste0(
  "const GENE_PANELS = {",
  paste(vapply(names(all_p), function(nm) {
    genes_json = paste0('"', all_p[[nm]], '"', collapse = ", ")
    paste0('"', nm, '": new Set([', genes_json, '])')
  }, character(1)), collapse = ",\n"),
  "};\n",
  'const DEFAULT_PANEL = "', params$default_panel, '";\n'
)
cat("<script>\n", js_panels, "\n</script>\n", sep = "")
```

## `r params$sample_id` {.unnumbered}

```{r header-badges}
mode_badge_col = if (outputs$mode == "matched") "#28a745" else "#fd7e14"
badges = tagList(
  tags$span(class = "badge",
            style = paste0("background-color:", mode_badge_col, "; color:white; margin-right:6px; padding:4px 8px; border-radius:4px;"),
            toupper(outputs$mode)),
  tags$span(class = "badge",
            style = "background-color:#6c757d; color:white; margin-right:6px; padding:4px 8px; border-radius:4px;",
            toupper(params$reference)),
  tags$span(class = "badge",
            style = "background-color:#6c757d; color:white; padding:4px 8px; border-radius:4px;",
            toupper(params$sex))
)
badges
```

```{r summary-cards}
fmt_val = function(x, digits = 2) {
  if (is.na(x)) return("N/A")
  if (is.numeric(x) && !is.integer(x)) return(round(x, digits))
  as.character(x)
}

card = function(label, value, colour = "#007bff") {
  div(
    style = paste0("display:inline-block; margin:6px; padding:12px 20px; ",
                   "border-left:4px solid ", colour, "; background:#f8f9fa; ",
                   "border-radius:4px; min-width:140px;"),
    div(style = "font-size:0.75em; color:#6c757d; text-transform:uppercase;", label),
    div(style = "font-size:1.4em; font-weight:bold;", value)
  )
}

tagList(
  card("Purity",         fmt_val(ascat_pp$purity),           "#d9534f"),
  card("Ploidy",         fmt_val(ascat_pp$ploidy),           "#f0ad4e"),
  card("Mean coverage",  paste0(fmt_val(qc_mosdepth$mean_depth), "×"), "#5bc0de"),
  card("Read N50",       if (!is.na(qc_cramino$n50)) fmt_bp(qc_cramino$n50) else "N/A", "#5cb85c"),
  card("Somatic SNVs",   fmt_val(n_snv, 0),                  "#9467bd"),
  card("Somatic SVs",    fmt_val(n_sv, 0),                   "#8c564b"),
  card("Panel variants", fmt_val(n_panel_vars, 0),            "#e377c2"),
  card("Panel SVs",      fmt_val(n_panel_svs, 0),             "#17becf")
)
```

---

## Circos overview

> Tracks (outer → inner): **ideogram**, **somatic SNVs** (coloured by 6-class SBS category), **non-BND SVs** (DEL/DUP/INV/INS), **ASCAT copy number** (major = red, minor = blue, total = black). Translocation links shown inside.

```{r circos-plot, fig.align='center'}
if (!is.null(circos_tmp) && file.exists(circos_tmp)) {
  img_b64 = base64enc::base64encode(circos_tmp)
  tags$img(src  = paste0("data:image/svg+xml;base64,", img_b64),
           style = "max-width:750px; display:block; margin:auto;")
} else {
  tags$div(class = "alert alert-warning",
           "Circos plot could not be generated. Check that ASCAT and Severus output files are present.")
}
```

---

## Small variants in gene panel

```{r gene-filter-ui, results='asis'}
panel_opts = ""
all_p = if (!is.null(params$all_panels) && length(params$all_panels) > 0)
          params$all_panels else list()
for (nm in names(all_p)) {
  selected = if (nm == params$default_panel) " selected" else ""
  label = paste0(toupper(substr(nm, 1, 1)), substr(nm, 2, nchar(nm)))
  panel_opts = paste0(panel_opts,
    '<option value="', nm, '"', selected, '>', label, '</option>\n')
}
panel_opts = paste0(panel_opts,
  '<option value="__all__">All genes</option>\n',
  '<option value="__custom__">Custom&hellip;</option>\n')

cat(paste0('
<div style="margin-bottom:12px;">
  <label for="panel-select" style="font-weight:600; margin-right:8px;">Gene panel:</label>
  <select id="panel-select" style="padding:4px 8px; border-radius:4px; border:1px solid #ced4da;">',
    panel_opts,
  '</select>
  <span id="panel-variant-count" style="margin-left:12px; font-size:0.9em; color:#6c757d;"></span>
</div>
<div id="custom-gene-panel" style="display:none; margin-bottom:12px;">
  <label style="font-weight:600; display:block; margin-bottom:4px;">
    Custom genes (one per line, or comma/space separated, case-insensitive):
  </label>
  <textarea id="custom-gene-input"
    rows="4" style="width:400px; font-family:monospace; padding:6px;
                    border:1px solid #ced4da; border-radius:4px;"></textarea>
</div>
'))
```

```{r small-variant-table}
if (!is.null(variant_table) && nrow(variant_table) > 0) {
  # Format VAF columns as percentages
  vaf_cols = grep("^vaf_", names(variant_table), value = TRUE)
  dt_display = copy(variant_table)
  for (vc in vaf_cols) {
    dt_display[, (vc) := round(get(vc) * 100, 1)]
  }
  setnames(dt_display, vaf_cols,
           sub("vaf_", "VAF% ", vaf_cols))

  DT::datatable(
    dt_display,
    rownames    = FALSE,
    filter      = "top",
    elementId   = "snv-table",
    extensions  = c("Buttons", "Scroller"),
    options     = list(
      dom        = "Bfrtip",
      buttons    = c("copy", "csv", "excel"),
      scrollX    = TRUE,
      scrollY    = "400px",
      scroller   = TRUE,
      deferRender = TRUE,
      pageLength = 25,
      columnDefs = list(list(className = "dt-left", targets = "_all")),
      initComplete = JS("function() { window.snvTableElem = this.api().table().node(); }")
    )
  ) |>
    DT::formatStyle(
      columns     = "impact",
      target      = "cell",
      backgroundColor = DT::styleEqual(
        c("HIGH", "MODERATE", "LOW", "MODIFIER"),
        c("#ffcccc", "#fff3cc", "#e6ffcc", "#f5f5f5")
      )
    )
}
```

---

## Structural variants in gene panel

```{r sv-info}
if (is.null(sv_table) || nrow(sv_table) == 0) {
  tags$div(class = "alert alert-info",
    if (is.null(sev_tsv)) "Severus gene-annotated SV file not found."
    else "No structural variants detected.")
}
```

```{r sv-table}
if (!is.null(sv_table) && nrow(sv_table) > 0) {
  DT::datatable(
    sv_table,
    rownames   = FALSE,
    filter     = "top",
    elementId  = "sv-table",
    extensions = c("Buttons", "Scroller"),
    options    = list(
      dom       = "Bfrtip",
      buttons   = c("copy", "csv", "excel"),
      scrollX   = TRUE,
      scrollY   = "350px",
      scroller  = TRUE,
      deferRender = TRUE,
      pageLength = 25,
      initComplete = JS("function() { window.svTableElem = this.api().table().node(); }")
    )
  ) |>
    DT::formatStyle(
      columns = "SVTYPE",
      target  = "cell",
      backgroundColor = DT::styleEqual(
        c("DEL", "DUP", "INV", "INS", "BND"),
        c("#cce5ff", "#d4edda", "#fff3cd", "#f8d7da", "#e2e3e5")
      )
    )
}
```

---

## QC details

::: {.callout-note collapse="true"}
### Coverage summary

```{r coverage-table}
if (!is.null(qc_mosdepth$table) && nrow(qc_mosdepth$table) > 0) {
  DT::datatable(qc_mosdepth$table, rownames = FALSE,
                options = list(pageLength = 30, dom = "ft", scrollY = "300px"))
} else {
  "Mosdepth summary not available."
}
```
:::

::: {.callout-note collapse="true"}
### Alignment statistics (samtools flagstat)

```{r flagstat-table}
if (length(qc_flagstat) > 0) {
  fs = data.frame(metric = names(qc_flagstat), value = unlist(qc_flagstat))
  knitr::kable(fs, row.names = FALSE)
} else {
  "Flagstat file not available."
}
```
:::

::: {.callout-note collapse="true"}
### Read quality (cramino)

```{r cramino-table}
cr = qc_cramino
cr_df = data.frame(
  metric = c("N50 (bp)", "Yield (Gb)", "% mapped", "# reads"),
  value  = c(fmt_val(cr$n50, 0), fmt_val(cr$yield_gb), fmt_val(cr$mapped_pct), fmt_val(cr$n_reads, 0))
)
knitr::kable(cr_df, row.names = FALSE)
```
:::

::: {.callout-note collapse="true"}
### Wakhan (haplotype-resolved CN)

```{r wakhan-status}
if (outputs$has_wakhan) {
  "Wakhan output directory found. Full Wakhan integration is planned for v2 of this report."
} else {
  "Wakhan was not run for this sample."
}
```
:::

---

*Report generated `r format(Sys.time(), "%Y-%m-%d %H:%M")` · LRSomatic report v1.0*

```{=html}
<script>
(function () {
  let activePanel = DEFAULT_PANEL;
  let customGenes = new Set();

  // Cache the NHL_GENE_HITS column index for the SV table (looked up lazily on first use)
  let svNhlColIdx = -1;
  function getSvNhlColIdx() {
    if (svNhlColIdx >= 0) return svNhlColIdx;
    if (!window.svTableElem) return -1;
    $(window.svTableElem).find('thead th').each(function (i) {
      if ($(this).text().trim() === 'NHL_GENE_HITS') { svNhlColIdx = i; }
    });
    return svNhlColIdx;
  }

  $.fn.dataTable.ext.search.push(function (settings, data) {
    const node = settings.nTable;
    if (node !== window.snvTableElem && node !== window.svTableElem) return true;
    if (activePanel === "__all__") return true;

    let hits;
    if (node === window.snvTableElem) {
      hits = [(data[0] || "").toUpperCase()];
    } else {
      const idx = getSvNhlColIdx();
      if (idx < 0) return true;
      hits = (data[idx] || "").toUpperCase().split(/[;,]+/).map(function (g) { return g.trim(); }).filter(function (g) { return g.length > 0; });
    }

    if (activePanel === "__custom__") return hits.some(function (g) { return customGenes.has(g); });
    const panel = GENE_PANELS[activePanel];
    return panel ? hits.some(function (g) { return panel.has(g); }) : true;
  });

  function parseCustomGenes(text) {
    return new Set(
      text.split(/[\s,\n]+/).map(g => g.trim().toUpperCase()).filter(g => g.length > 0)
    );
  }

  function updateCount() {
    const snvTbl = window.snvTableElem ? $(window.snvTableElem).DataTable() : null;
    if (!snvTbl) return;
    const n = snvTbl.rows({ search: "applied" }).count();
    const el = document.getElementById("panel-variant-count");
    if (el) el.textContent = n + " SNV" + (n === 1 ? "" : "s") + " shown";
  }

  function redraw() {
    if (window.snvTableElem) $(window.snvTableElem).DataTable().draw();
    if (window.svTableElem)  $(window.svTableElem).DataTable().draw();
    updateCount();
  }

  $(document).ready(function () {
    $("#panel-select").on("change", function () {
      activePanel = this.value;
      const isCustom = activePanel === "__custom__";
      $("#custom-gene-panel").toggle(isCustom);
      redraw();
    });

    let debounceTimer;
    $("#custom-gene-input").on("input", function () {
      clearTimeout(debounceTimer);
      debounceTimer = setTimeout(function () {
        customGenes = parseCustomGenes(document.getElementById("custom-gene-input").value);
        redraw();
      }, 300);
    });

    let _waitAttempts = 0;
    const waitForTables = setInterval(function () {
      if (window.snvTableElem || window.svTableElem) {
        clearInterval(waitForTables);
        redraw();
      } else if (++_waitAttempts > 100) {
        clearInterval(waitForTables);
      }
    }, 100);
  });
})();
</script>
```
